Efficiently stable presentations from error-correcting codes

### 2023-10-26 efficient_stability.tex
Updates from Overleaf

\paragraph{Acknowledgments.} 
MC acknowledges with gratitude the Simons Society of Fellows and is supported by a grant from the Simons Foundation (N. 965535). 
TV is supported by a research grant from the Center for New Scientists at the Weizmann Institute of Science, a Simons Investigator award, AFOSR Grant No. FA9550-22-1-0391, and ERC Consolidator Grant VerNisQDevS (101086733). HY is supported by AFOSR award FA9550-21-1-0040, NSF CAREER award CCF-2144219, and the Sloan Foundation.

@@

For $q$ a prime power we let $\F_q$ denote the finite field with $q$ elements. 
For $n,k,d$ integer, a $[n,k,d]_q$ linear code $\code$ is a $k$-dimensional subspace of $\F_q^n$ such that for all $x\in \code$ such that $x\neq 0$, the Hamming weight $|x|$ (i.e.\ the number of nonzero coordinates of $x$) is at least $d$. The parameter $n$ is called the \emph{length} of the code, $k$ its \emph{dimension} and $d$ its \emph{distance}. A code can be specified by a \emph{parity check matrix} $h\in \F_q^{m\times n}$ such that $\code = \ker h$.

@@

\begin{proof}
The group defined by the right-hand side is obviously Abelian and a $2$-group, so it is of the form $\Z_2^{k'}$ for some $k'$. In fact, it is equal to the quotient of $\Z_2^n$ by the subgroup generated by the $\prod_{1\leq j \leq n} x_j^{h_{ij}}$. So it is $\Z_2^k$ where $k=n-\dim\textrm{im}\ h = \dim\ker h=\dim \code$.  
\end{proof}

@@

An $r$-local linear tester $M=(h,\nu)$ for $\code$ induces a pair of distributions $(\nu_R,\nu_S)$ on the relations and generators of the presentation $G(h)$\eqref{eq:def-gh-pres} in a natural way: For the generators, we let $\nu_S$ be induced from $\nu$ by first sampling $j\sim \nu$ and then a uniformly random $i$ such that $h_{ji}\neq 0$. For the relations, we let $\nu_R$ be the uniform mixture of the distribution $\nu_S$ on relations $x_j^2=e$, the distribution $\nu$ on relations $R_i$, and the distribution $\nu\times \nu_S\times \nu_S$ on relations $R'_{jii'}$.

@@

Let $q=2^t$ and $\code$ an $[n,k,d]_q$ linear code. Let $\code_{\Had}$ be the Hadamard code over $\F_2^t$ (introduced at the end of Section~\ref{sec:pres-code}). 
Let $\code'$ be the $[qn,tk,d']$ linear code over $\F_2$ defined as follows. Given $a\in (\F_2^t)^{k}$, first map $a\mapsto a'=\kappa^{-1}(a) \in \F_q^{k}$. Then encode $a'$ to $b'=\code(a')\in \F_q^n$. Finally, return $b=\code_\Had(\kappa(b'))\in(\F_2^q)^n$, where $\code_\Had$ is applied component-wise. Using that $\code_\Had$ has distance $q/2$, it is easy to verify that this code has distance $d'\geq dq/2$.

Given an $r$-local tester $M=(h,\nu)$ for $\code$, there is a natural $\max(r,3)$-local tester $M'$ for $\code'$ which can be described as follows. Index coordinates of $\code'$ by pairs $(i,\alpha)\in [n]\times\F_2^t$, fixing a bijection between $[qn]$ and $[n]\times \F_2^t$.  We describe an $\max(r,3)$-local \mnote{shouldn't this be max(r,3)-local}\tnote{yes thanks fixed} tester $M' = (h',\nu')$ for $\code'$. Informally, $h'$ contains two type of checks. First, the $A$-checks consist in the repetition of $n$ copies of the checks for the Hadamard code, one for each Hadamard-code encoding of an $\F_q$-symbol from $\code$. Second, the $B$-checks implement the checks of $\code$ specified by $M$, directly on the Hadamard encoding. More precisely, define $h'$ to be the block matrix $h'=\begin{pmatrix} A \\ B \end{pmatrix}$ where 
\begin{itemize}
	\item $A \in \F^{nq^2 \times nq}$ is itself a block-diagonal matrix where the diagonal blocks are the $q^2 \times q$ parity check matrix for the Hadamard code. 
	In other words, $A$ can be viewed as $I_{n\times n} \otimes h_\had$ where $h_\had \in \F_2^{q^2 \times q}$ is the parity check matrix for the Hadamard code. 
	\item $B \in \F^{\ell q \times nq}$ is viewed as having rows indexed by pairs $(p,\gamma) \in \{1,\ldots,\ell\} \times \F_q$ and columns indexed by pairs $(i,x) \in \{1,\ldots,n\} \times \F_2^t$. The entry in row $(p,\gamma)$ and column $(i,x)$ is $1$ if and only if $h_{pi} \neq 0$ and $x = \kappa(\gamma h_{pi})$.
\end{itemize}
Define the distribution $\nu'$ as the uniform mixture of the uniform distribution on the rows of the $A$ block matrix and the uniform distribution on the rows of the $B$ block matrix.

@@

\begin{proof}
Let $x\in \F_2^{qn}$ be such that $h'x=0$. Since the $A$-checks enforce that each block of $k$ symbols contains the Hadamard-code encoding of an $\F_q$ symbol, $x$ can be decoded to $x'\in \F_q^n$ such that for each $(i,x)\in \{1,\ldots,n\}\times \F_2^t$, $x_{(i,x)} = \kappa(x'_i)\cdot x$. If the $p$-th row of $h$ enforces the check $v_p\cdot x'=0$, where $v_p\in \F_q^n$, then 
the $(p,\gamma)$-th row of $B$ enforces the check 
\begin{align*}
0&=\sum_{j=1}^n \kappa(\gamma (v_p)_j) \cdot \kappa(x'_j)\\
&= \sum_{j=1}^n \tr( \gamma (v_p)_j x'_j) \\
&= \tr(\gamma (v_p\cdot  x')\;.
\end{align*}
Therefore, $h'x=0$ is equivalent to $\tr(\gamma(v_p \cdot x'))=0$ for all $\gamma$, which is equivalent to $v_p\cdot x'=0$. Thus $\code'=\ker h'$, as desired. 
\end{proof}

>>>>

\paragraph{Acknowledgments.} 
We would like to thank John Wright for his remarks on an early draft of this paper. MC acknowledges with gratitude the Simons Society of Fellows and is supported by a grant from the Simons Foundation (N. 965535). 
TV is supported by a research grant from the Center for New Scientists at the Weizmann Institute of Science, a Simons Investigator award, AFOSR Grant No. FA9550-22-1-0391, and ERC Consolidator Grant VerNisQDevS (101086733). HY is supported by AFOSR award FA9550-21-1-0040, NSF CAREER award CCF-2144219, and the Sloan Foundation.

@@

For $q$ a prime power we let $\F_q$ denote the finite field with $q$ elements. 
For $n,k,d$ integer, an $[n,k,d]_q$ linear code $\code$ is a $k$-dimensional subspace of $\F_q^n$ such that for all $x\in \code$ such that $x\neq 0$, the Hamming weight $|x|$ (i.e.\ the number of nonzero coordinates of $x$) is at least $d$. The parameter $n$ is called the \emph{length} of the code, $k$ its \emph{dimension} and $d$ its \emph{distance}. A code can be specified by a \emph{parity check matrix} $h\in \F_q^{m\times n}$ such that $\code = \ker h$.

@@

\begin{proof}
The group defined by the right-hand side is obviously abelian and a $2$-group, so it is of the form $\Z_2^{k'}$ for some $k'$. In fact, it is equal to the quotient of $\Z_2^n$ by the subgroup generated by the $\prod_{1\leq j \leq n} x_j^{h_{ij}}$. So it is $\Z_2^k$ where $k=n-\dim\textrm{im}\ h = \dim\ker h=\dim \code$.  
\end{proof}

@@

An $r$-local linear tester $M=(h,\nu)$ for $\code$ induces a pair of distributions $(\nu_R,\nu_S)$ on the relations and generators of the presentation $G(h)$~\eqref{eq:def-gh-pres} in a natural way: For the generators, we let $\nu_S$ be induced from $\nu$ by first sampling $j\sim \nu$ and then a uniformly random $i$ such that $h_{ji}\neq 0$. For the relations, we let $\nu_R$ be the uniform mixture of the distribution $\nu_S$ on relations $x_j^2=e$, the distribution $\nu$ on relations $R_i$, and the distribution $\nu\times \nu_S\times \nu_S$ on relations $R'_{jii'}$.

@@

Let $q=2^t$ and $\code$ be an $[n,k,d]_q$ linear code. Let $\code_{\Had}$ be the Hadamard code over $\F_2^t$ (introduced at the end of Section~\ref{sec:pres-code}). 
Let $\code'$ be the $[qn,tk,d']$ linear code over $\F_2$ defined as follows. Given $a\in (\F_2^t)^{k}$, first map $a\mapsto a'=\kappa^{-1}(a) \in \F_q^{k}$. Then encode $a'$ to $b'=\code(a')\in \F_q^n$. Finally, return $b=\code_\Had(\kappa(b'))\in(\F_2^q)^n$, where $\code_\Had$ is applied component-wise. Using that $\code_\Had$ has distance $q/2$, it is easy to verify that this code has distance $d'\geq dq/2$.

Given an $r$-local tester $M=(h,\nu)$ for $\code$, there is a natural $\max(r,3)$-local tester $M'$ for $\code'$ which can be described as follows. Index coordinates of $\code'$ by pairs $(i,\alpha)\in [n]\times\F_2^t$, fixing a bijection between $[qn]$ and $[n]\times \F_2^t$.  We describe an $\max(r,3)$-local \mnote{shouldn't this be max(r,3)-local}\tnote{yes thanks fixed} tester $M' = (h',\nu')$ for $\code'$. Informally, $h'$ contains two type of checks. First, the $A$-checks consist of the repetition of $n$ copies of the checks for the Hadamard code, one for each Hadamard-code encoding of an $\F_q$-symbol from $\code$. Second, the $B$-checks implement the checks of $\code$ specified by $M$, directly on the Hadamard encoding. More precisely, define $h'$ to be the block matrix $h'=\begin{pmatrix} A \\ B \end{pmatrix}$ where 
\begin{itemize}
	\item $A \in \F^{nq^2 \times nq}$ is itself a block-diagonal matrix where the diagonal blocks are the $q^2 \times q$ parity check matrix for the Hadamard code. 
	In other words, $A$ can be viewed as $I_{n\times n} \otimes h_\had$ where $h_\had \in \F_2^{q^2 \times q}$ is the parity check matrix for the Hadamard code. 
	\item $B \in \F^{\ell q \times nq}$ is viewed as having rows indexed by pairs $(p,\gamma) \in \{1,\ldots,\ell\} \times \F_q$ and columns indexed by pairs $(i,x) \in \{1,\ldots,n\} \times \F_2^t$. The entry in row $(p,\gamma)$ and column $(i,x)$ is $1$ if and only if $h_{pi} \neq 0$ and $x = \kappa(\gamma h_{pi})$.
\end{itemize}
Define the distribution $\nu'$ as the uniform mixture of the uniform distribution on the rows of the $A$ block matrix and the uniform distribution on the rows of the $B$ block matrix.

@@

\begin{proof}
Let $x\in \F_2^{qn}$ be such that $h'x=0$. Since the $A$-checks enforce that each block of $k$ symbols contains the Hadamard-code encoding of an $\F_q$ symbol, $x$ can be decoded to $x'\in \F_q^n$ such that for each $(i,x)\in \{1,\ldots,n\}\times \F_2^t$, $x_{(i,x)} = \kappa(x'_i)\cdot x$. If the $p$-th row of $h$ enforces the check $v_p\cdot x'=0$, where $v_p\in \F_q^n$, then 
the $(p,\gamma)$-th row of $B$ enforces the check 
\begin{align*}
0&=\sum_{j=1}^n \kappa(\gamma (v_p)_j) \cdot \kappa(x'_j)\\
&= \sum_{j=1}^n \tr( \gamma (v_p)_j x'_j) \\
&= \tr(\gamma (v_p\cdot  x'))\;.
\end{align*}
Therefore, $h'x=0$ is equivalent to $\tr(\gamma(v_p \cdot x'))=0$ for all $\gamma$, which is equivalent to $v_p\cdot x'=0$. Thus $\code'=\ker h'$, as desired. 
\end{proof}
